How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-07-08 doc/response_review.tex

\documentclass[12pt,letterpaper]{article}

\usepackage{amsmath, amsthm, amsfonts, amssymb}
\usepackage{microtype, parskip, graphicx}
%\usepackage[comma,numbers,sort&compress]{natbib}
\usepackage[style=numeric-comp,
      backend=biber,
      bibencoding=utf8,
      maxbibnames=10,
      uniquename=init,
      giveninits=true,
      sorting=none,
      natbib=true,
      url=false,
      doi=false,
      isbn=false]{biblatex}
\usepackage[margin=1in]{geometry}
\usepackage{lineno}
\usepackage{longtable}
\usepackage{caption, subcaption, multirow, morefloats, rotating}
\usepackage{wrapfig}
\usepackage{hyperref}
\usepackage{threeparttable}

\frenchspacing

% https://3d.bk.tudelft.nl/hledoux/blog/fiddling-biblatex/
%-- formatting hell for biblatex
%-- remove "In:"
\renewbibmacro{in:}{}
%-- no "quotes" around titles of chapters/article titles
\DeclareFieldFormat[article, inbook, incollection, inproceedings, misc, thesis, unpublished]
{title}{#1}
%-- no punctuation after volume
\DeclareFieldFormat[article]{volume}{ {#1} } 
%-- puts number/issue between brackets
\DeclareFieldFormat[article, inbook, incollection, inproceedings, misc, thesis, unpublished]{number}{\mkbibparens{#1}} 
%-- and then for articles directly the pages w/o any "pages" or "pp." 
\DeclareFieldFormat[article]{pages}{#1}
%-- for some types replace "pages" by "p."
\DeclareFieldFormat[inproceedings, incollection, inbook]{pages}{p. #1}

% https://tex.stackexchange.com/questions/81569/biblatex-parentheses-around-the-volume-number-of-an-article
\renewbibmacro*{volume+number+eid}{%
 \printfield{volume}%
 \printfield{number}%
 \printfield{eid}}
\DeclareFieldFormat[article]{number}{\mkbibparens{#1}}

% bibliographic information
%\bibliographystyle{abbrvnat}
%\bibliography{citations}
\addbibresource{citations.bib}

\title{Responses to reviews of ``How predictable is extinction? Forecasting species survival at million-year timescales''}
\author{
 Smits, Peter\\
 %Department of Integrative Biology, University of California Berkeley\\
 \texttt{psmits@berkeley.edu} 
 \and
 Finnegan, Seth\\
 %Department of Integrative Biology, University of California Berkeley\\
 \texttt{sethf@berkeley.edu}
}
\date{}

\begin{document}

\begin{refsection}

\maketitle

\linenumbers{}
\modulolinenumbers[3]

Below are our responses to the editor and reviewer comments that we recieved on our original manuscrpt. We would like to thank the editor and the three reviewers for their thoughtful and insightful comments which have greatly improved the manuscript.

Editor and reviewer comments are presented in bold. Below these comments, we detail how we updated our manuscript in response to that comment.

\section{Editor comments}

\begin{bfseries}
  Line 5 (abstract): comma after ‘to answer this question’

  Line 7: period after parenthesis

  Line 10: comma before ‘the improvement in predictive power’

  Line 25: ‘the suggestion’ or ‘suggestions’
\end{bfseries}

We have updated the relevant sections of the text with these suggestions.

\begin{bfseries}
  Line 31-33: I think this sentence needs to be clarified. Can you describe more clearly what is meant by ‘how accurate or strong our predictions about the differences in extinction risk are’, and how this differs from strong and general determinants of extinction risk?
\end{bfseries}

In our general reorganization of the introduction section of this paper we tried to improve the clarity of our language. We thank the editor for pointing out this sentence as it was redundant and confusing.

\begin{bfseries}
  Line 34: try to avoid ending a sentence with a preposition

  Line 40: among taxa, rather than between?

  Line 41: comma before ‘and the relative risk of…’

  Line 77: need space before ‘Occurrences were filtered…’

  Line 78: ‘was no more’…is that right?
\end{bfseries}

We have updated the relevant sections of the text with these suggestions.

\begin{bfseries}
  Line 83: what about taxa with age ranges that were more uncertain than 1 mys? Did this ever occur? [Ah, I read further and now see your call to the SI]
\end{bfseries}

Thank you for pointing this issue out. We have tried to improve our references to the supplemental material by putting more references in the text instead of just a single reference at the end of the section.

\begin{bfseries}
  Line 86: I echo Wolfgang’s comment in wondering how congruent would be patterns if they would be measured using another range-based metric?
\end{bfseries}

Thank you both for the suggestion to move to minimum spanning tree measure of geographic range. We have updated our analysis using this new range metric. The change in range metric did not change our interpretation of the results.

\begin{bfseries}
  Line 115: insert ‘in’ prior to ‘Section S1.3’
\end{bfseries}

We have updated the relevant section of the text with this suggestion.

\begin{bfseries}
  Line 114(ish): have you considered testing change in geographic range over multiple intervals? I.e., maybe a stronger signal would be observed if there was a persistent decline in range size over multiple 1 my time bins? Similarly, species may be buffered from extinction if their range increases over multiple time bins. It would be interesting to tease whether this is more important than absolute size…
\end{bfseries}

Thank you for this suggestion. We have update our analysis to include change in geographic range from the last three observations. This change adds two more covariates to all models that include our ``historical covariates.'' This addition to our analysis did change our results, and we now consider our most complex model that includes historical covariates and variation in covariate effect over time to be overfit to our data. Our models that include either historical covariates or variation in effects over time, however, are found to have functionally equivalent out-of-sample performance.

\begin{bfseries}
  Line 119: try to avoid ending a sentence with a preposition

  Line 183: ‘Comparing’?

  Line 195: I am probably being dense, but how can the in-sample have a poorer performance than the out-of-sample? Isn’t this fairly counter-intuitive?

  Line 214: ‘forecast’ vs ‘forecasting’?

  Line 224: ‘an’ before ‘extremely important…’
\end{bfseries}

We have updated the relevant section of the text with this suggestion, in particular focusing on clarity.

\section{Reviewer comments}

\subsection{Reviewer 1}

\begin{bfseries}
  Smits and Finnegan use the Neptune database to check which combination of potentially influencing variables has the best predictive power of extinction risk in the marine plankton. They use sophisticated logistic modelling of the time-binned raw data to suggest that considering changes of parameters only marginally improves prediction accuracy. Overall, this is a very well written, although difficult to comprehend, account, which is to the point and rich in insights.

  While the main result is well founded by the sophisticated analyses (in fact too sophisticated to be understandable in all details by the reviewer), it is somewhat disappointing, as the relative importance of parameters in multivariate models remain vague. It is also disappointing to see that change of range had a negligible effect in the multivariate context, as this is one of the main arguments for using the fossil record in extinction risk assessments (besides taxon age and intrinsic turnover rates). I argue that, to really discard change of range, the authors would need to look at longer trajectories of range. Bin-to-bin changes at this temporal resolution (1 myr) are perhaps too noisy to inform models. I am not asking for additional analyses for this paper, but this point should be added to the discussion.
\end{bfseries}

Thank you for this suggestion. We have update our analysis to include change in geographic range from the last three observations. This change adds two more covariates to all models that include our ``historical covaraites.'' This addition to our analysis did change our results, and we now consider our most complex model that includes historical covariates and variation in covariate effect over time to be overfit to our data. Our models that include either historical covariates or variation in effects over time, however, are found to have functionaly equivalent out-of-sample performance.

\begin{bfseries}
  We are confronted with four models and even the simplest one is already quite complex. Is there a chance to provide for each model the relative importance of each variable for the model performance? Alternatively, as standing geographic range is among the best predictors of extinction risk in paleo studies, can the authors add a very simple model of just event ~ range? I’d be curious to see if the AUC is then “significantly” lower than in the other models. Of course, in a frequentist world, I would start with a complex model and then perform model selection with a step function. I’m not Bayesian enough to suggest a specific strategy for this paper, but somehow model performance needs to be balanced by model complexity. The authors seem to have compared just model performance.
\end{bfseries}

Thank you for pointing out this issue in our text and we apologize for the lack of clarity. We have updated our methods section to further explain cross-validation and how we have implicitly dealt with model complexity because of the linked definitions of cross-validation and AIC. From our updated text:

``Cross-validation is a procedure for estimating a models expected out-of-sample error. Information criteria such as AIC or WAIC are approximations of out-of-sample predictive error as estimated by cross-validation \citep{ESL,Gelman2013d}. Cross-validation implicitly takes into account model complexity because when a model is overfit to its data, out-of-sample predictions will be biased and inaccurate \citep{ESL}. A high degree of similarity between out-of-sample and in-sample estimates indicates that the model is not overfit to the data (though it is not necessarily an adequate descriptor of the data). Cross-validation is preferable to simple metrics such as AIC because instead of a single value it produces, an entire posterior distribution of estimates.''

\begin{bfseries}
  The methods are very well explained (especially in the supplement), although uncertainties in parameter estimates merit some further prose. Using maximum great circle distance (GCD) as a measure of geographic range is ok but prone to sampling bias. Better sampled intervals will inevitably have greater mean ranges than more poorly sampled intervals. While the scaling applied by the authors takes different means and variance into account it may not be sufficient. If GCD is used, I’d recommend normalizing to the maximum observed geographic range in an interval prior to scaling. Grid cell occupancy or minimum spanning trees are alternative measures, which are worthy of exploration. 
\end{bfseries}

Thank you both for the suggestion to move to minimum spanning tree measure of geographic range. We have updated our analysis using this new range metric. The change in range metric did not change our interpretation of the results.

We chose not to standardize geographic range to in-bin max because we believe this breaks ``coherence'' of the measure and the interpretation of the regression coefficients. With geographic range not standardized by max in-bin geographic range, the regression coefficients for the ``change in range'' covariates are the expected change in log-odds of extinction risk per standard deviation change in geographic range between that previous bin and the current bin. In contrast, if the variables are defined as percentages, the difference between bins gives a less obvious or interpretable regression coefficient. Additionally, max in-bin geographic range is nearly identical for the entire Cenozoic.

\begin{bfseries}
  Looking at change of range is good, but the authors could also look at changes beyond two intervals. Although models incorporating multiple bins may result in overfitting, the argument is that long-term decline will increase extinction risk. I’d argue that two bins (especially at the relatively fine temporal scale applied) can hardly measure long-term decline. Finally, the authors should provide a rationale why these and not other parameters were used in the study (see also below). Does latitude or skeletal mineralogy have little to add to model performance?
\end{bfseries}

Thank you for this suggestion. We have update our analysis to include change in geographic range from the last three observations. This change adds two more covariates to all models that include our ``historical covariates.'' This addition to our analysis did change our results, and we now consider our most complex model that includes historical covariates and variation in covariate effect over time to be overfit to our data. Our models that include either historical covariates or variation in effects over time, however, are found to have functionally equivalent out-of-sample performance.

We have modified our discussion sections to further explain the rational behind our covariate choices. Simply, our goal is to provide a baseline for extinction prediction that can be compared to more specific models which include these more specific covariates. Additionally, a lot of the suggested covariates are either not available for a lot of the taxa throughout the Cenozoic or are nearly constant for the entire taxonomic group. In many ways, by allowing our parameter estimates vary by taxonomic group we have attempted to implictly deal with these types of ecological differences.

\begin{bfseries}
  Presentation of results could be improved. Showing raw data of extinction rates in the four groups, and individual analyses how age and phyla actually influence extinction risk would be preferred over the wiggly lines in Figs. 2 and 3. The latter could be moved to the supplement. The text in the results section is very technical, describing too much detail. 
\end{bfseries}

Thank you for this suggestion. We have added a section to our supplemental material that includes figures presenting the parameter estimates for the group-level regression coefficients and the individual-level estimates for those regression coefficients over time.

\begin{bfseries}
  The discussion is good, but may have to be partly rewritten in light of my comments.

  Specific comments (line numbers refer to the smaller ones on the right):

  - “the relative risk of extinction exhibited by different taxonomic groups” could be abbreviated as “intrinsic extinction risk”

  - l. 68-69. Partial repetition of sentences before. “Incredible” should be omitted. If incredible why trust the age models?

  - l. 79. “Ecologically unusual” should be omitted from reasoning. Not needed. 
\end{bfseries}

We have updated the relevant sections of the text with these suggestions.

\begin{bfseries}
  - l. 88-91. Authors may want to add that Mg/Ca has its own issues, especially the steep rise of Mg/Ca in seawater through the Cenozoic. Cramer et al. have discussed and accounted for this trend, but with some assumptions, which result in considerable uncertainties.
\end{bfseries}

Thank you for pointing this out. We have updated our Methods section with a discussion of the potential problems with Mg/Ca based temperature estimates.

\begin{bfseries}
  - l. 132: paragraph starts with k-fold but is specified for 5-fold. I recommend rewriting this paragraph for the general application and then specify.

  - l. 149-150: “so we interpret values between 0.7 and 0.8 could then be considered”  “so we consider values between   0.7 and 0.8  as”

  - Caption Fig. 1: Cut “a” from “with a higher AUC values”. It is not entirely clear what the error bars and bean plots represent. Are the AUC values derived from the 62 time bins or is there more to it? Please explain.

  - l. 173: “Dinoflagellates” should be “diatoms” unless I have missed something. Same in labels of Figs. 2-4 and some incidences in the text. Please correct throughout.

  - Xlab of Fig. 2 should be “Mya” or similar instead of “CI”

  - l. 176-177. Repetition from methods.

  - l. 183. “Comparison”  “Comparing”

  - l. 192-193. No single-sentence paragraphs

  - l. 197. “diatoms” instead of “Dinoflagellates”

  - l. 214-216. To my understanding AUC does not simply measure a rank-order analysis but model performance. That the AUC’s of the eight models are statistically indistinguishable (are they?) does not necessarily imply a correct forecasting of rank-order extinction probability.  Please rephrase!
\end{bfseries}

We have updated the relevant sections of the text and our figures with these suggestions.

\begin{bfseries}
  - l. 223-224. “Variation of extinction intensity”. Would be useful to either depict this variation or report some summary statistics (median, range, variance).
\end{bfseries}

Thank you for this suggestion. We have added a section to our supplmental material that includes figures presenting the parameter estimates for the group-level regression coefficients and the individual-level estimates for those regression coefficients over time. These estimates include average log-odds of extinction over time.

\begin{bfseries}
  - l. 233. What does “scientifically significant” mean is this context? I am aware that simple model selection is not applicable here but it would be good to have a comparable metric.
\end{bfseries}

Thank you for pointing out this lack of clarity. We have removed this section and attempted to improve the clarity of our discussion of our results.

\begin{bfseries}
  - l. 236-237. Kiessling and Kocsis and used more than just corals. Their coarser temporal resolution (geological stages) is an obvious suspect for the greater role of change of occupancy than reported here. 
\end{bfseries}

We have update our covarites to include additional lags. Please see the updated Methods, Results, and Discussion for how our results have changed.

\begin{bfseries}
  - l. 244-245. Poor reasoning. For example skeletal mineralogy (silica versus calcium carbonate), trophic level (phyto- vs. zooplankton), latitudinal preference (e.g., median paleolatitude) or additional physico-chemical parameters (e.g., sea-level, Mg/Ca, pCO2) are intuitive variables, whereas taxon-age is less intuitive in light of the recent literature. I’m not asking for additional analyses or more complex models but a sound rationale.
\end{bfseries}

We have modified our discussion sections to further explain the rational behind our covariate choices. Simply, our goal is to provide a baseline for extinction prediction that can be compared to more specific models which include these more specific covariates. Additionally, a lot of the suggested covariates are either not available for a lot of the taxa throughout the Cenozoic or are nearly constant for the entire taxonomic group. In many ways, by allowing our parameter estimates vary by taxonomic group we have attempted to implictly deal with these types of ecological differences.

\begin{bfseries}
  - l. 247. Add “for” between “not exist” and “all”. In any case, a poor defense given the obvious traits to look at (see above).

  - l. 254. Modern “conservation determinations” are not continuous but ordinal. The difference between “critically endangered” and “endangered” may be very different from the difference between “endangered” and “vulnerable”.

  - l. 255-256. I argue that the authors’ results actually show that fossil data are not very relevant. If this impression is wrong, the authors should perhaps rephrase some parts of the text.

  Also check use of “My” vs. “Mya” and cases in plankton groups. As a general rule these should be upper case in a formal context (e.g., Radiolaria) and lower case when used informally (e.g., radiolarians). 
\end{bfseries}

We have updated the relevant sections of the text with these suggestions.

\begin{bfseries}
  The supplement is good but lacks sensitivity tests and a rationale of why the specific parameters and not others where chosen in the models. 
\end{bfseries}

Thank you for this suggestion. We have improved the language in our methods section surrounding in-sample verus out-of-sample forecasts. Our in-sample forecasts are a type of sensitivity test.

From the updated manuscript:

``In-sample forecasting is a posterior predictive check in that we are estimating our model's ability to correctly classify the data to which it was fit. Posterior predictive checks are a type of sensitivity analysis because we are checking the quality of model's fit to the data. If our models have poor in-sample forecasting performance, than our models are not adequate descriptors of the data and will most likely make poor out-of-sample predictions. In-sample forecasting measures, however, are not the same as understanding our models' ability to forecast future extinctions or if our models are overfit to our data and produce biased out-of-sample estimates \citep{ESL}.''

And:

``The in-sample model comparisons are useful for comparing the relative ability of our models to represent the data they were fit to, acting as quality control and sensitivity analysis. Comparison between the posterior distributions \dots''

\subsection{Reviewer 2}

\begin{bfseries}
  This manuscript uses a high-resolution record of marine microfossils to test the predictive power of fossil record extinctions. A lot of paleontological research has focused on determining the importance of particular biological and environmental risk factors, so this study is novel in testing whether past history can predict future extinction. The methodology is explained well and the main result is well-supported.

  I only have two broader thoughts and two more specific notes:

  I find it interesting that the predictability of extinction doesn’t obviously differ between background intervals and intervals with more unusual environmental perturbations (such as the PETM or E-O transition). Perhaps that’s because those events weren’t mass extinctions in the vein of the P/T or K/Pg. There’s also a fair amount of volatility in the AUC time-series. Nevertheless, does this consistent predictive ability regardless of differences in conditions say anything interesting about extinction predictions?

### 2019-07-08 doc/article.tex
process for resubmit

Past palaeobiological studies of extinction have frequently focused on identifying and measuring the effect of various potential predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Ezard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}. This focus means that while we have a good understanding of which factors are strong and general determinants of extinction risk, we have less knowledge of how capable we are at predicting future extinction events.

We ask how accurate risk predictions based on fossil data are when estimating future extinctions. Because future extinctions are unobservable we cannot directly evaluate the ultimate performance of such predictions. However, we can address the question by taking a given point in the geological past, developing a predictive model based on extinction patterns prior to that point, and assessing the predictive performance of this model on unobserved (e.g. ``future'', from the point of view of the model) extinction/survival events.

Extinction intensity (average rate) and selectivity (difference in risk among taxa) vary through time, and the relative risk of extinction exhibited by different taxonomic groups and how that risk varies over time is an important dynamic which shapes the rate and structure of extinction \citep{Payne2007,Payne2016,Ezard2011,Smits2019}. This variation also raises a question: given that extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future? Putting aside the important question of how human activities will alter the determinants of future extinction risk, we can address this uncertainty by specifically including and modeling the temporal variation in extinction risk across a range of extinction intensities and selectivities.

Numerous studies have established that geographic range is one of the most important determinants of extinction risk in the fossil record, and that a species geographic range can be highly variable over geologic time \citep{Foote2007,Liow2010,Liow2007,Kiessling2013,Payne2007,Jablonski2003,Jablonski2008,Jablonski2006}. The degree to which the past can help to predict the future fates of species depends in part on the degree to which species’ geographic range trajectories are deterministic versus Markovian \citep{Liow2007,Kiessling2013,Foote2007b,Pigot2012}. In the former case, knowledge of the specific past trajectory of a species -- whether its range has expanded or contracted over a given time span -- might help to improve assessments of its current risk. In the latter case, only the current range of the species conveys useful information about its current risk, although we can still use prior extinction patterns to augment predictions by evaluating the relative extinction risk of species that had similar ranges in the past. Discriminating among these alternative models is thus very important for determining how best to incorporate fossil data in present risk assessments.

For this exercise we chose to analyze one of the best-sampled and studied fossil records -- the Cenozoic record of skeletonized marine planktonic microorganisms (Foraminifera, Radiolaria, Diatoms, and Coccolithophores). These data are readily available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project \citep{Lazarus1994,SpencerCervato1999}. This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an taxonomy synonymization strategy for delaing wiht 50+ years of taxonomic opinion \citep{Lazarus1994} -- as close to ideal data for this analysis as possible. Analyzing patterns of extinction and global occurrence at fine temporal scales means we can better elucidate how well we can predict species extinction at human-relevant scales.

@@

We analyzed microfossil occurrence information from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}. This occurrence-based dataset includes calcareous nannoplankton, diatoms, planktonic foraminifera, and radiolarians. Occurrences were filtered to include only those species who's first occurrence is no earlier than 63 Mya (millions of years ago). This filtering criterion excludes taxa that survived the K/Pg extinction or arose during this recovery interval, and ensures that our occurrence histories fully overlap with the temperature time-series used as a potential extinction risk predictor (see below).

All fossil occurrences were assigned to 1 My (million year) bins based on the estimated age of the fossil occurrence as listed in the Neptune Database. After binning, each taxon's geographic range was calculated for each of the 1 My bins in which it occurred. Geographic range was calculated as the minimum spanning tree distance between all observations of that taxon during that temporal bin; this distance was measured in kilometers. Minimum spanning tree distance was caluclated using the \texttt{GeoRange} package for R \citep{GeoRange}.

We also included how a taxon's geographic range has changed since its last three observation times. We measured this change in geographic range by calculating the difference in geographic range between an observation and that taxon's three previous occurrences. If there are not enough previous observations to calculate, then that value is recorded as a 0. These differences were calculated after minimum spanning tree distance was transformed and standardized (see Section \ref{sec:data_desc}).

@@

We also included the global temperature from the previous time that taxon was observed. If there are not enough previous observations to calculate, then that value is recorded as a 0. This lag was calculated after global temperature was transformed and standardized (see Section \ref{sec:data_desc}).

@@

See Section \ref{sec:data_desc} for a further explanation on how observations were temporally binned, and how our covariates were standardized and transformed prior to analysis.

@@

Similarly, we included time of obseration as an additional varying-intercept term to account for changes average global extinction risk over time that are not related to the covariates included in this model. This varying-intercept is further allowed to vary by taxonomic group. This varying-intercept term allows us to tease apart the differences in extinction risk associated with time of observation versus age since first observation. An important note is that for our V and VP models, the covariation between this varying-intercept and the varying-slopes of our covariates is explicitly modeled (see Section \ref{sec:model_desc}).

See Table \ref{tab:model_def} for further explanation of how the four models we considered differ from each other. A complete description of the statistical model used in this analysis is available in Section \ref{sec:model_desc}. Additionally, the full description of how these models were implemented and coded, including choice of priors, is available in Section \ref{sec:model_est}.

>>>>

Extinction intensity (average rate) and selectivity (difference in risk among taxa) have varied greatly through time, and the relative risk of extinction exhibited by different taxonomic and ecological groups can provide insights into the drivers of both background and mass extinction \citep{Payne2007,Payne2016,Ezard2011,Smits2019,Wang2008a,Knoll2007}. Many studies have examined the effects of various potential predictors on extinction risk through time \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Harnik2012a,Ezard2011,Foote2006} or refined methods for identifying and measuring these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}. These studies have produced a growing body of knowledge regarding which factors have been general determinates of extinction risk in the geological past.

A related question that has received much less attention is how successful we might expect to be when using this knowledge to attempt to predict future extinction events.

Because future extinctions are unobservable we cannot directly evaluate the ultimate performance of such predictions. However, we can take a given point in the geological past, develop a predictive model based on extinction patterns prior to that point, and assessing the predictive performance of this model on subsequent (e.g. ``future'', from the point of view of the model) extinction/survival events. Putting aside the very important question of how human activities will alter the determinants of future extinction risk, such an approach provides a framework for evaluating the expected accuracy of future risk assessments based on past extinction events.

%This variation also raises a question: given that extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future? Putting aside the important question of how human activities will alter the determinants of future extinction risk, we can address this uncertainty by specifically including and modeling the temporal variation in extinction risk across a range of extinction intensities and selectivities.

Here we take this approach, using as a model system the Cenozoic record of skeletonized marine planktonic microorganisms (Foraminifera, Radiolaria, Diatoms, and Coccolithophores). This record has several key strengths for our purposes: planktonic microorganisms are widespread and abundant in pelagic habitats, have high preservation potential, and because of their utility for biostratigraphic, paleoclimatic, and oceanographic study they have been the focus of an extensive international coring and study effort \citep{Lazarus1994,SpencerCervato1999}. A compliation of these data is readily available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project \citep{Lazarus1994,SpencerCervato1999}. This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an taxonomic synonymization framework for dealing with 50+ years of taxonomic opinion \citep{Lazarus1994} -- as close to ideal data for this analysis as possible. Analyzing patterns of extinction and global occurrence at fine temporal scales means we can better elucidate how well we can predict species extinction at human-relevant scales.

The overall question of how well models based on past extinction patterns perform at forecasting future extinctions depends in part on model complexity. Simple models requiring only a few parameters are in general preferable because more complex models run a greater risk of being overfit to the observations on which they are trained. In addition, many traits that might influence extinction risk among extant species are difficult to assign confidently to extinct species. For these reasons we elect to focus on “baseline” models which include only a few parameters that have been shown to be important and/or consistent determinants of extinction risk in the marine fossil record. Numerous studies have established that geographic range is one of the most important determinants of extinction risk in the fossil record, and that a species’ geographic range can be highly variable over geologic time \citep{Foote2007,Liow2010,Liow2007,Kiessling2013,Payne2007,Jablonski2003,Jablonski2008,Jablonski2006}. In addition to geographic range, we also considered global climate state and change in climate state since previous observation in order to evaluate the influence of climate or climate change trajectory on extinction risk. Finally, we included species age, both because previous studies of planktonic taxa have found it to be a determinant of extinction risk and because its inclusion in our models is critical to their nature as survival models (see Model Specifications below). We reiterate that our primary objective is to evaluate the predictive performance of simple models that include only a few general parameters; more complex models including other likely determinants of extinction risk such as skeletal mineralogy, trophic ecology, and thermal tolerance range might well perform better.

There are a number of ways in which past extinction patterns might be used to model present risk. The simplest case assumes that relationships between predictors (hereafter covariates) and extinction risk have been constant through time. A more complex but more realistic case allows relationships between covariates and extinction risk to vary through time, consistent with evidence for temporal variation in extinction selectivity regime. Finally, an important consideration is the degree to which species’ geographic range trajectories exhibit deterministic versus Markovian behavior \citep{Liow2007,Kiessling2013,Foote2007b,Pigot2012}. In the former case, knowledge of the specific past trajectory of a species – whether its range has expanded or contracted from some point in the past to the present – might help to improve assessments of its current risk. In the latter case only the current geographic range of the species would convey useful information about current and future risk (although that assessment would still be based on the relative extinction risk of species that had similar ranges at different points in the geological past).

Below, we evaluate four models along a spectrum from simplest (fixed covariate effects, Markovian range dynamics) to most complex (varying covariate effects, deterministic range dynamics). We ask (1) how well they perform at classifying species as extinctions or survivors in the data they were fit to, and (2) how well they perform at classifying species as extinctions or survivors in “future” data that were not used in fitting the models.

@@

We analyzed microfossil occurrence information from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}. This occurrence-based dataset includes calcareous nannoplankton, diatoms, planktonic foraminifera, and radiolarians. Occurrences were filtered to include only those species with first occurrences no earlier than 63 Mya (millions of years ago). This filtering criterion excludes taxa that survived the K/Pg extinction or arose during this recovery interval, and ensures that our occurrence histories fully overlap with the temperature time-series used as a potential extinction risk predictor (see below).

All fossil occurrences were assigned to 1 My (million year) bins based on the estimated age of the fossil occurrence as listed in the Neptune Database. After binning, each species' geographic range was calculated for each of the 1 My bins in which it occurred. Geographic range was calculated as the minimum spanning tree distance between all observations of that taxon during that temporal bin; this distance was measured in kilometers. Minimum spanning tree distance was calculated using the \texttt{GeoRange} package for R \citep{GeoRange}.

We also included how a taxon's geographic range has changed since its last three observation times. We measured this change in geographic range by calculating the difference in geographic range between an observation and that taxon's three previous occurrences. Change between the most recent and the three previous occurrences was calculated individually for each of those lags. If there are not enough previous observations to calculate, then that value is recorded as a 0. These differences were calculated after minimum spanning tree distance was transformed and standardized (see Supplement Section S1.1.2)

@@

We also included the global temperature from the previous time that taxon was observed. If there are not enough previous observations to calculate, then that value is recorded as a 0. This lag was calculated after global temperature was transformed and standardized (see Supplement Section S1.1.2).

@@

See Supplemental Section S1.1 for a further explanation on how observations were temporally binned, and how our covariates were standardized and transformed prior to analysis.

@@

Similarly, we included time of obseration as an additional varying-intercept term to account for changes average global extinction risk over time that are not related to the covariates included in this model. This varying-intercept is further allowed to vary by taxonomic group. This varying-intercept term allows us to tease apart the differences in extinction risk associated with time of observation versus age since first observation. An important note is that for our V and VP models, the covariation between this varying-intercept and the varying-slopes of our covariates is explicitly modeled (see Supplement Section S1.2).

See Table \ref{tab:model_def} for further explanation of how the four models we considered differ from each other. A complete description of the statistical model used in this analysis is available in Supplement Section S1.2. Additionally, the full description of how these models were implemented and coded, including choice of priors, is available in Supplement Section S1.2.

\begin{table}[ht]
 \caption{Models and their definitions}
 \begin{threeparttable}
  {
   \def\arraystretch{1.5}
   \begin{tabular}{ l p{3cm} l l }
    Code & Description & Covariates & R Formula Syntax\tnote{a}\phantom{\textsuperscript{a}} \\
    \hline
    C & Constant effects, no historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event\tnote{b}\phantom{\textsuperscript{b}} $\sim$ range\tnote{c}\phantom{\textsuperscript{c}} + temp\tnote{d}\phantom{\textsuperscript{d}} +\\(1 $|$ time\tnote{e}\phantom{\textsuperscript{e}}/phylum\tnote{f}\phantom{\textsuperscript{f}}) +\\(1 $|$ age\tnote{g}\phantom{\textsuperscript{g}}/phylum)} \\
    V & Varying effects, no historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + temp +\\(1 + range + temp $|$ time/phylum) + (1 $|$ age/phylum)} \\ 
    CP & Constant effects, historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range, temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ + range\_diff1\tnote{g}\phantom{\textsuperscript{h}} + range\_diff2\tnote{h}\phantom{\textsuperscript{h}} + range\_diff3\tnote{h}\phantom{\textsuperscript{h}} + \\temp + temp\_lag\tnote{i}\phantom{\textsuperscript{i}} +\\(1 $|$ time/phylum) +\\(1 $|$ age/phylum)} \\
    VP & Varying effects, historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range, temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + range\_diff1 + range\_diff2 + range\_diff3 + \\temp + temp\_lag +\\(1 + range + range\_diff1 + range\_diff2 + range\_diff3 +\\temp + temp\_lag $|$ time/phylum) +\\(1 $|$ age/phylum)} \\
    \hline
   \end{tabular}
  }
  \begin{tablenotes}
  \item[a] See Supplemental Equation S2 for full statistical model definition.
  \item[b] Species observation where 1 if time of last observation, otherwise 0.
  \item[c] Species geographic range in log km\(^2\). Mean centered, scaled to sd = 1.
  \item[d] Global temperature in degrees C. Mean centered, scaled to sd = 1.
  \item[e] Time of observation.
  \item[f] Taxonomic group of species (i.e. Foraminifera, Diatoms, Radiolarians, Calcareous nannoplankton).
  \item[g] Age at observation.
  \item[h] Change in geographic range since last observation (number indicates how lags).
  \item[i] Temperature at previous observation.
  \end{tablenotes}
 \end{threeparttable}
 \label{tab:model_def}
\end{table}
